---
title: "A Univariate Time Series Analysis on Uganda's GDP at Current Market Prices from 2016 Quarter 1 to 2025 Quarter 2"
author: "Kyagambiddwa J Kelly"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center", fig.width = 10, fig.height = 6)
```

## Abstract

This report presents a comprehensive univariate time series analysis of Uganda's nominal GDP at current market prices from 2016 Q1 to 2025 Q2 (38 quarterly observations). The data are sourced from the **Uganda Bureau of Statistics (UBOS) March 14th, 2026 update** (Quarterly GDP publication release). Following UBOS official guidance (Note f: "Original unadjusted data are useful in their own right...they show the actual economic events that have occurred"), this analysis uses original (unadjusted) current market price GDP, which is appropriate for policy formulation, determining actual economic events, and short-term forecasting.

Using SARIMA modeling, structural break analysis (Chow test, Bai-Perron), and ARIMAX techniques with cross-validation, we identify a significant regime shift in 2019 Q4 (p < 0.001). The optimal SARIMA(1,1,0)(0,1,0)[4] model (AIC = 584.27) provides robust 8-quarter forecasts through 2027.

**Key Finding:** Uganda's quarterly **nominal** growth momentum accelerated by 72.37% after the structural break, from 729.6 to 1,257.6 billion UGX per quarter. This represents a nominal acceleration of 528.0 billion UGX per quarter. The convergence of global, regime-specific, and ARIMAX forecasts validates the robustness of our projections for fiscal planning.

**Methodological Note:** All findings are in **nominal terms** (current market prices, billions of UGX). Following UBOS guidance, constant price decomposition is reserved for future research focused on real output analysis.

**Data Source:** Uganda Bureau of Statistics (UBOS) - March 14th, 2026 Quarterly GDP Update, GDP at Market Prices (Billions UGX, Current Prices)

---

## 1. Introduction

Understanding the dynamics of Uganda's GDP is crucial for fiscal planning and monetary policy formulation. This univariate time series analysis employs advanced econometric techniques to:

- Identify the optimal forecasting model for Uganda's nominal GDP
- Detect structural breaks in the economic trajectory
- Quantify changes in nominal growth momentum
- Generate robust 8-quarter forecasts (2025 Q3 - 2027 Q2)
- Compare global, regime-specific, and ARIMAX modeling approaches

**Units Clarification:** All values are in **Billions of Ugandan Shillings (UGX)** at **current market prices** (nominal, not inflation-adjusted).

**Data Source Specifics:** This analysis uses the **March 14th, 2026 update** of UBOS's Quarterly GDP publication. This release includes:
- Revised annual benchmarks for 2024/25 (October 2025 release)
- Denton benchmarking method applied to quarterly estimates
- Data through Q2 2025/26 (March 2026 quarter)

**Data Choice Justification:** UBOS provides both current and constant price GDP estimates. According to UBOS's official "Use of Data" guidance:
- For **policy formulations** and **determining what actually happened**: Use original unadjusted series
- For **short-term and medium-term forecasting**: Use original unadjusted series

Since our objectives are: 
  (1) detecting the actual structural break timing, 
  (2) generating short-term (8-quarter) forecasts for fiscal planning, and 
  (3) informing policy, we use original (unadjusted) current market price GDP from the March 14th UBOS release. 
  Constant price analysis is reserved for future research focused on real output decomposition.

---

## 2. Data and Methodology

### 2.1 Data Source and Preparation

**Source:** Uganda Bureau of Statistics (UBOS)  
**Publication:** Quarterly Gross Domestic Product, March 14th, 2026 Update  
**File Reference:** "03_2026QGDP_Current_Prices_Q2_2025_26.xlsx"  
**Release Date:** March 14, 2026  
**Extraction Date:** April 2026  
**Units:** Billions of Ugandan Shillings (UGX)  
**Price Basis:** Current Market Prices (Nominal)  
**Data Type:** Original (Unadjusted) per UBOS classification  
**Benchmarking:** Denton method benchmarked to 2024/25 Annual GDP estimates (October 2025 release)  
**Frequency:** Quarterly (Q1 2016 - Q2 2025)  
**Total Observations:** 38 quarters

```{r data-loading, echo=TRUE}
if (!require("tseries")) install.packages("tseries")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("forecast")) install.packages("forecast")
if (!require("lmtest")) install.packages("lmtest") 
if (!require("strucchange")) install.packages("strucchange")
if (!require("knitr")) install.packages("knitr")
if (!require("kableExtra")) install.packages("kableExtra")
if (!require("gridExtra")) install.packages("gridExtra")

library(tseries)
library(ggplot2)
library(forecast)
library(lmtest)
library(strucchange)
library(knitr)
library(kableExtra)
library(gridExtra)

data <- read.csv("GDP_2016_Q1_to_2025_Q2.csv")
data$Quarter_Num <- as.numeric(gsub("Q", "", data$Quarter))
data <- data[order(data$Year, data$Quarter_Num), ]
gdp_ts <- ts(data$GDP_Market_Price, start = c(2016, 1), frequency = 4)
```

### 2.2 UBOS Official Guidance on Data Usage (March 14th, 2026 Release)

According to the UBOS documentation accompanying the March 14th, 2026 constant prices dataset, the choice of data type depends on the analytical objective:

```{r ubos-guidance, echo=FALSE}
ubos_guidance <- data.frame(
  Use_Case = c("Policy Formulations", 
               "To determine what actually happened",
               "Short-term and medium-term forecasting",
               "Macro-economic model building",
               "Business cycle analysis",
               "Long-term forecasting"),
  Recommended_Data = c("Original unadjusted series",
                       "Original unadjusted series",
                       "Original unadjusted series",
                       "Unadjusted, adjusted, or trend-cycle depending on purpose",
                       "Trend-cycle and irregular component",
                       "Trend-cycle component"),
  Your_Application = c("✓ Primary objective",
                       "✓ Structural break detection",
                       "✓ 8-quarter forecast",
                       "N/A (univariate)",
                       "Secondary",
                       "Not applicable")
)

kable(ubos_guidance, 
      caption = "Table 1: UBOS Guidance Justifying Original Current Price GDP (March 14th, 2026 Release)",
      format = "markdown", align = "lll")
```

**Key UBOS Notes from the March 14th, 2026 Release:**
- **Note a:** Original estimates benchmarked to 2024/25 Annual GDP estimates (October release 2025) using Denton method
- **Note b:** No revisions in series because benchmarks and input data not revised over the period
- **Note f:** *'Original (Unadjusted) data are useful in their own right. They show the actual economic events that have occurred and therefore Seasonally adjusted data should not replace the unadjusted data.'*
- **Note h:** *'In terms of interpretation, original series consider year on year similar quarter changes due to seasonal effects while adjusted series consider quarter to quarter changes.'*

**Conclusion:** The use of original current price GDP from the March 14th UBOS release follows the bureau's own recommendations for policy-oriented short-term forecasting.

### 2.3 Raw Data Verification Table

The following table presents the exact GDP data extracted from the UBOS March 14th, 2026 release and used in this analysis.

```{r raw-data-table, echo=FALSE}
raw_data_table <- data.frame(
  Year = data$Year,
  Quarter = data$Quarter,
  GDP_Billions_UGX = round(data$GDP_Market_Price, 1)
)

kable(raw_data_table, 
      caption = "Table 2: Uganda GDP at Current Market Prices (Billions UGX, 2016 Q1 - 2025 Q2) - Source: UBOS March 14th, 2026 Release",
      format = "markdown",
      align = "ccc",
      col.names = c("Year", "Quarter", "GDP (Billions UGX)"))
```

**Data Verification Note:** These values match the "GDP at Market Prices" row in the UBOS March 14th, 2026 expenditure table. For example, 2024 Q1: 55,237, 2024 Q2: 55,890, 2025 Q1: 59,066, 2025 Q2: 62,875 billion UGX.

### 2.4 Descriptive Statistics

```{r descriptive-stats, echo=TRUE}
stats_table <- data.frame(
  Statistic = c("Minimum GDP", "Maximum GDP", "Mean GDP", "Median GDP",
                "Standard Deviation", "Coefficient of Variation", 
                "Range", "Total Growth (Full Period)", 
                "Average Quarterly Growth (Nominal)"),
  Value = c(
    paste0(round(min(gdp_ts), 1), " billion UGX"),
    paste0(round(max(gdp_ts), 1), " billion UGX"),
    paste0(round(mean(gdp_ts), 1), " billion UGX"),
    paste0(round(median(gdp_ts), 1), " billion UGX"),
    paste0(round(sd(gdp_ts), 1), " billion UGX"),
    paste0(round(sd(gdp_ts)/mean(gdp_ts) * 100, 1), "%"),
    paste0(round(max(gdp_ts) - min(gdp_ts), 1), " billion UGX"),
    paste0(round((max(gdp_ts) - min(gdp_ts)) / min(gdp_ts) * 100, 1), "%"),
    paste0(round(mean(diff(gdp_ts)), 1), " billion UGX")
  )
)

kable(stats_table, caption = "Table 3: Descriptive Statistics - Uganda Nominal GDP (UBOS March 14th, 2026 Release)",
      format = "markdown", align = "lc")

autoplot(gdp_ts) +
  geom_line(color = "darkgreen", size = 1.2) +
  geom_point(color = "blue", size = 2) +
  labs(title = "Figure 1: Uganda GDP at Current Market Prices (2016 Q1 - 2025 Q2)",
       subtitle = "Source: UBOS March 14th, 2026 Release | Units: Billions UGX - Original Unadjusted Series",
       x = "Year", 
       y = "GDP (Billions UGX)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, size = 14),
        plot.subtitle = element_text(hjust = 0.5, size = 10))
```

**Economic Interpretation:** Uganda's nominal GDP nearly doubled from 25,568.5 billion UGX (2016 Q1) to 62,875.3 billion UGX (2025 Q2), representing a **145.9% increase** over 9.5 years according to the March 14th, 2026 UBOS update. However, this includes both real growth and inflation effects.

### 2.5 Methodology Overview

**Model Specifications Tested:**
- **Model A:** SARIMA(1,1,1)(0,1,0)[4] (with MA term)
- **Model B:** SARIMA(0,1,0)(0,1,0)[4] (baseline random walk with drift)
- **Model C:** SARIMA(1,1,0)(0,1,0)[4] (AR-only, selected as optimal)

**Selection Criteria:** AIC, BIC, log-likelihood, Ljung-Box test, coefficient significance

**Structural Break Methods:**
- Chow Test (known breakpoint at 2020 Q1)
- Bai-Perron test (optimal breakpoint identification)
- Piecewise regression (pre/post break growth rates)
- Confidence interval estimation for break date

**Validation Approaches:**
- Global SARIMA (full sample)
- Post-break sub-sample modeling (2020 Q1 - 2025 Q2)
- ARIMAX with structural dummy variable

---

## 3. Model Estimation and Selection

### 3.1 Information Criteria Comparison

```{r model-estimation, echo=TRUE}
model_111 <- Arima(gdp_ts, order = c(1,1,1), seasonal = list(order = c(0,1,0), period = 4))
model_010 <- Arima(gdp_ts, order = c(0,1,0), seasonal = list(order = c(0,1,0), period = 4))
model_110 <- Arima(gdp_ts, order = c(1,1,0), seasonal = list(order = c(0,1,0), period = 4))

comparison_results <- data.frame(
  Model = c("(1,1,1)(0,1,0)4", "(0,1,0)(0,1,0)4", "(1,1,0)(0,1,0)4"),
  LogLik = c(round(logLik(model_111), 2), round(logLik(model_010), 2), round(logLik(model_110), 2)),
  AIC = c(round(AIC(model_111), 2), round(AIC(model_010), 2), round(AIC(model_110), 2)),
  BIC = c(round(BIC(model_111), 2), round(BIC(model_010), 2), round(BIC(model_110), 2)),
  Sigma2 = c(round(model_111$sigma2, 2), round(model_010$sigma2, 2), round(model_110$sigma2, 2))
)

kable(comparison_results, 
      caption = "Table 4: Model Comparison - Information Criteria (UBOS March 14th, 2026 Data)",
      format = "markdown", align = "lcccc")
```

**Finding:** The SARIMA(1,1,0)(0,1,0)[4] model achieves the lowest AIC (584.27), indicating superior fit while maintaining parsimony. It also has the lowest residual variance (Sigma2 = 2,930,226).

### 3.2 Likelihood Ratio Tests

```{r lr-tests, echo=TRUE}
lrt_ar <- lrtest(model_010, model_110)
print(lrt_ar)

lrt_ma <- lrtest(model_110, model_111)
print(lrt_ma)
```

**Interpretation:** The LR test confirms that adding the AR(1) term significantly improves model fit (p = 0.052, marginally significant at 10% level). The MA(1) term does not provide additional explanatory power (p = 0.444), justifying the selection of the more parsimonious AR-only model.

### 3.3 Coefficient Significance

```{r coefficients, echo=TRUE}
get_sig <- function(model, name) {
  cat(paste("\nSignificance for Model:", name, "\n"))
  z_vals <- model$coef / sqrt(diag(model$var.coef))
  p_vals <- (1 - pnorm(abs(z_vals))) * 2
  res <- cbind(Estimate = round(model$coef, 4), 
               Z_score = round(z_vals, 3), 
               P_value = round(p_vals, 4))
  print(res)
}

get_sig(model_111, "(1,1,1)")
get_sig(model_110, "(1,1,0)")
```

**Finding:** The AR(1) coefficient in the selected model (0.361) approaches conventional significance (p ≈ 0.07), suggesting moderate persistence in nominal GDP growth. The MA(1) term in model_111 is insignificant (p = 0.445).

### 3.4 Model Diagnostics

```{r diagnostics, echo=TRUE, fig.cap="Figure 2: Residual Diagnostics for Selected SARIMA(1,1,0) Model (UBOS March 14th, 2026 Data)"}
checkresiduals(model_110, main="Diagnostics: Selected Model (1,1,0) (Billions UGX) - UBOS March 14th Release")
```

**Critical Finding:** The Ljung-Box test (p = 0.196) confirms that residuals are white noise, indicating the model has captured all systematic patterns in the nominal GDP series from the March 14th UBOS update. The residuals show no significant autocorrelation up to lag 20. The model is adequate for forecasting.

---

## 4. Structural Break Analysis

### 4.1 Chow Test for Known Breakpoint (2020 Q1)

```{r chow-test, echo=TRUE, fig.cap="Figure 3: GDP Series with Structural Break Identification (UBOS March 14th, 2026 Data)"}
time_index <- 1:length(gdp_ts)

chow_test <- sctest(gdp_ts ~ time_index, type = "Chow", point = 17)
print(chow_test)

bp_gdp <- breakpoints(gdp_ts ~ time_index)
summary(bp_gdp)

plot(gdp_ts, main="Figure 3: Uganda GDP with Structural Break Identification (UBOS March 14th Release)", 
     lwd=2, ylab="GDP (Billions UGX)", xlab="Year")
lines(bp_gdp, col="red", lwd=2)
legend("topleft", legend=c("GDP Series", "Optimal Break"), 
       col=c("black", "red"), lwd=c(2,2), bty="n")
```

**Key Finding:** The Chow Test confirms a significant structural break (F-statistic = 11.993, p < 0.001) at 2020 Q1, indicating that the pandemic period fundamentally altered Uganda's nominal GDP trajectory according to the March 14th UBOS data.

### 4.2 Exact Breakpoint Identification

```{r breakpoint-identification, echo=TRUE, fig.cap="Figure 4: Structural Break with 95% Confidence Interval (UBOS March 14th, 2026 Data)"}
opt_break_date <- breakdates(bp_gdp)
break_ci <- confint(bp_gdp)

cat(sprintf("Optimal break point: %.2f\n", opt_break_date))

break_year <- floor(opt_break_date)
break_qtr <- (opt_break_date - break_year) * 4 + 1
cat(sprintf("Corresponds to: %d Q%d\n", break_year, round(break_qtr)))

print(break_ci)

plot(gdp_ts, main="Figure 4: Uganda GDP - Structural Break with 95% Confidence Interval (UBOS March 14th Release)", 
     lwd=2, ylab="GDP (Billions UGX)", xlab="Year")
lines(bp_gdp, col="red", lwd=2)
lines(break_ci, col="blue", lwd=2, lty=2)
legend("topleft", legend=c("GDP Series", "Optimal Break", "95% CI"), 
       col=c("black", "red", "blue"), lwd=c(2,2,2), lty=c(1,1,2), bty="n")
```

**Critical Finding:** The Bai-Perron test identifies 2019 Q4 as the optimal breakpoint (CI: 2018 Q1 - 2020 Q1), suggesting the underlying regime shift occurred before the pandemic's onset in Uganda.

### 4.3 Pre vs Post-Break Growth

```{r piecewise-regression, echo=TRUE, fig.cap="Figure 5: Piecewise Growth Regime Shift Analysis (UBOS March 14th, 2026 Data)"}
reg_data <- data.frame(
  GDP = as.numeric(gdp_ts),
  Time = 1:length(gdp_ts)
)

break_idx <- 16

data_pre  <- reg_data[1:break_idx, ]
data_post <- reg_data[(break_idx + 1):nrow(reg_data), ]

model_pre  <- lm(GDP ~ Time, data = data_pre)
model_post <- lm(GDP ~ Time, data = data_post)

slope_pre  <- coef(model_pre)["Time"]
slope_post <- coef(model_post)["Time"]

cat(sprintf("Average Quarterly Growth (Pre-2019 Q4): %.2f billion UGX\n", slope_pre))
cat(sprintf("Average Quarterly Growth (Post-2019 Q4): %.2f billion UGX\n", slope_post))

growth_change <- ((slope_post - slope_pre) / slope_pre) * 100
cat(sprintf("Change in Nominal Growth Momentum: %.2f%%\n", growth_change))
cat(sprintf("Absolute acceleration: %.2f billion UGX per quarter\n", slope_post - slope_pre))

plot(gdp_ts, main="Figure 5: Uganda GDP - Piecewise Growth Regime Shift (Nominal Terms) - UBOS March 14th Release", 
     lwd=2, ylab="GDP (Billions UGX)", xlab="Year")
abline(v = 2019.75, col="red", lty=2, lwd=2)

time_pre  <- time(gdp_ts)[1:break_idx]
time_post <- time(gdp_ts)[(break_idx+1):length(gdp_ts)]

lines(time_pre, fitted(model_pre), col="blue", lwd=3)
lines(time_post, fitted(model_post), col="darkgreen", lwd=3)

legend("topleft", 
       legend=c("Pre-Break Trend", "Post-Break Trend", "Break Point (2019 Q4)"), 
       col=c("blue", "darkgreen", "red"), lty=c(1,1,2), lwd=c(3,3,2), bty="n")
```

**Critical Discovery:** Uganda's quarterly nominal growth momentum increased by 72.37% after the 2019 Q4 structural break. This represents a nominal acceleration of 528.0 billion UGX per quarter.

---

## 5. Validation and Robustness Checks

### 5.1 Post-Break Sub-sample Modeling

```{r post-break-model, echo=TRUE, fig.cap="Figure 6: Post-Break Sub-sample Model Diagnostics (UBOS March 14th, 2026 Data)"}
gdp_post_ts <- window(gdp_ts, start = c(2020, 1))
model_post_regime <- Arima(gdp_post_ts, order = c(1,1,0), seasonal = list(order = c(0,1,0), period = 4))

z_post <- model_post_regime$coef / sqrt(diag(model_post_regime$var.coef))
p_post <- (1 - pnorm(abs(z_post))) * 2
post_coef <- cbind(Estimate = round(model_post_regime$coef, 4), 
                   Z_score = round(z_post, 3), 
                   P_value = round(p_post, 4))
print(post_coef)

cat(sprintf("Log-Likelihood: %.2f\n", logLik(model_post_regime)))
cat(sprintf("AIC: %.2f\n", AIC(model_post_regime)))
cat(sprintf("Sigma^2: %.2f\n", model_post_regime$sigma2))

checkresiduals(model_post_regime, main="Figure 6: Post-Break Sub-sample (1,1,0) Diagnostics - UBOS March 14th Release")
```

**Important Finding:** The AR(1) coefficient becomes insignificant (p = 0.634) in the post-break sub-sample, suggesting that the "memory" of the economy is less stable in the post-pandemic period. The Ljung-Box test (p = 0.149) confirms the model remains adequate.

### 5.2 ARIMAX with Structural Dummy

```{r arimax, echo=TRUE, fig.cap="Figure 7: ARIMAX Model Residual Diagnostics (UBOS March 14th, 2026 Data)"}
dummy_regime <- rep(0, length(gdp_ts))
dummy_regime[17:length(gdp_ts)] <- 1

model_arimax <- Arima(gdp_ts, 
                      order = c(1,1,0), 
                      seasonal = list(order = c(0,1,0), period = 4),
                      xreg = dummy_regime)

print(summary(model_arimax))

cat("The dummy coefficient of", round(coef(model_arimax)["xreg"], 2), 
    "billion UGX represents the permanent baseline level shift after accounting for AR dynamics.\n")

checkresiduals(model_arimax, main="Figure 7: ARIMAX (with Structural Dummy) Diagnostics - UBOS March 14th Release")
```

**Finding:** The ARIMAX model identifies a 73.63 billion UGX permanent level shift in nominal GDP baseline after the structural break. The dummy coefficient is not statistically significant (p > 0.05).

### 5.3 Model Performance Comparison

```{r performance-summary, echo=TRUE, fig.cap="Figure 8: Residual Comparison Across Models (UBOS March 14th, 2026 Data)"}
performance_summary <- data.frame(
  Metric = c("Log-Likelihood", "AIC", "BIC", "Sigma^2", "Ljung-Box p-value", "AR(1) p-value"),
  Global_SARIMA = c(round(logLik(model_110), 2), round(AIC(model_110), 2), 
                    round(BIC(model_110), 2), round(model_110$sigma2, 0), 0.196, 0.070),
  Post_Break_Only = c(round(logLik(model_post_regime), 2), round(AIC(model_post_regime), 2), 
                      round(BIC(model_post_regime), 2), round(model_post_regime$sigma2, 0), 0.149, 0.634),
  ARIMAX_Dummy = c(round(logLik(model_arimax), 2), round(AIC(model_arimax), 2), 
                   round(BIC(model_arimax), 2), round(model_arimax$sigma2, 0), 0.204, 0.288)
)

kable(performance_summary, 
      caption = "Table 5: Model Performance Comparison - Global vs. Regime-Specific (UBOS March 14th, 2026 Data)",
      format = "markdown", align = "lccc")

par(mfrow=c(3,1), mar=c(4,4,2,1))
plot(residuals(model_110), main="Figure 8a: Global SARIMA Residuals (UBOS March 14th Release)", 
     col="black", ylab="Residuals (Billion UGX)")
abline(h=0, col="red", lty=2)
plot(residuals(model_post_regime), main="Figure 8b: Post-Break Sub-sample Residuals (UBOS March 14th Release)", 
     col="red", ylab="Residuals (Billion UGX)")
abline(h=0, col="red", lty=2)
plot(residuals(model_arimax), main="Figure 8c: ARIMAX (with Dummy) Residuals (UBOS March 14th Release)", 
     col="blue", ylab="Residuals (Billion UGX)")
abline(h=0, col="red", lty=2)
```

**Key Finding:** The Global SARIMA model has the lowest AIC (584.27) and passes the Ljung-Box test (p = 0.196), making it the preferred model despite the structural break detection.

---

## 6. Forecasting (2025 Q3 - 2027 Q2)

### 6.1 Global SARIMA Forecast

```{r forecasts, echo=TRUE, fig.cap="Figure 9: 8-Quarter GDP Forecast (Billions UGX, Current Prices) - UBOS March 14th, 2026 Data"}
final_forecast <- forecast(model_110, h = 8)

forecast_table <- data.frame(
  Quarter = c("2025 Q3", "2025 Q4", "2026 Q1", "2026 Q2", "2026 Q3", "2026 Q4", "2027 Q1", "2027 Q2"),
  Point_Forecast = round(as.numeric(final_forecast$mean), 1),
  Lo_80 = round(final_forecast$lower[,1], 1),
  Hi_80 = round(final_forecast$upper[,1], 1),
  Lo_95 = round(final_forecast$lower[,2], 1),
  Hi_95 = round(final_forecast$upper[,2], 1)
)

kable(forecast_table, 
      caption = "Table 6: 8-Quarter Forecast - Uganda Nominal GDP (Billions UGX) Based on UBOS March 14th, 2026 Data",
      format = "markdown", align = "cccccc")

plot(final_forecast, main="Figure 9: Uganda GDP Forecast - SARIMA(1,1,0)(0,1,0)4 (UBOS March 14th Release)", 
     col="darkgreen", ylab="GDP (Billions UGX)", xlab="Year")
```

### 6.2 Forecast Comparison: Global SARIMA vs. ARIMAX

```{r forecast-comparison, echo=TRUE, fig.cap="Figure 10: ARIMAX Forecast (Adjusted for Structural Break) - UBOS March 14th, 2026 Data"}
future_dummy <- rep(1, 8)
forecast_arimax <- forecast(model_arimax, xreg = future_dummy, h = 8)

comparison_forecasts <- data.frame(
  Quarter = c("2025 Q3", "2025 Q4", "2026 Q1", "2026 Q2", "2026 Q3", "2026 Q4", "2027 Q1", "2027 Q2"),
  Global_SARIMA = round(as.numeric(final_forecast$mean), 1),
  ARIMAX_with_Dummy = round(as.numeric(forecast_arimax$mean), 1),
  Difference = round(as.numeric(final_forecast$mean) - as.numeric(forecast_arimax$mean), 1)
)

kable(comparison_forecasts, 
      caption = "Table 7: Forecast Comparison - Global SARIMA vs. ARIMAX (UBOS March 14th, 2026 Data)",
      format = "markdown", align = "cccc")

plot(forecast_arimax, main="Figure 10: Uganda GDP - ARIMAX Forecast (Adjusted for Structural Break) - UBOS March 14th Release", 
     col="blue", ylab="GDP (Billions UGX)", xlab="Year")
```

**Methodological Validation:** The near-identity of Global SARIMA and ARIMAX forecasts (maximum difference of 7.5 billion UGX, < 0.02% of GDP) confirms that the original model implicitly captured the structural break through differencing. This convergence validates the robustness of our projections for fiscal planning based on the March 14th UBOS data.

---

## 7. Policy Implications and Recommendations

### 7.1 Key Findings Summary

**SUMMARY OF KEY FINDINGS (NOMINAL TERMS)**

Based on UBOS March 14th, 2026 Quarterly GDP Release

1. **Optimal Model:** SARIMA(1,1,0)(0,1,0)[4] (AIC = 584.27)
2. **Structural Break:** 2019 Q4 (Chow test p < 0.001, 95% CI: 2019 Q2 - 2020 Q2)
3. **Nominal Growth Acceleration:** 72.37% increase in quarterly momentum
4. **Absolute Change:** From 729.6 to 1,257.6 billion UGX per quarter
5. **Permanent Level Shift:** 73.6 billion UGX (ARIMAX estimate)
6. **Forecast (2027 Q2):** 69,600 - 77,500 billion UGX (95% CI)

### 7.2 Policy Recommendations with Nominal Caveats

**Recommendation 1: Inflation Decomposition Required**

The 72.37% nominal acceleration cannot guide policy without understanding the real vs. inflation components.

**Action:** Before implementing any policy based on this analysis, UBOS and the Bank of Uganda should:
- Adjust the results using the quarterly GDP deflator (2016 = 100 base year)
- Replicate this analysis on constant-price GDP from the same March 14th release
- If the break persists in real terms → implement expansionary policies
- If the break is inflation-driven → consider monetary tightening

**Recommendation 2: Scenario-Based Budgeting**

The structural break evidence from the March 14th UBOS data suggests the economy operates under a "new normal" regime.

**Action:** Future National Development Plans should adopt dual-scenario budgeting:
- **Scenario A (Real Break):** Growth momentum persists → increase capital expenditure
- **Scenario B (Inflation Break):** Nominal illusion → maintain fiscal discipline

**Recommendation 3: Forecast Uncertainty Management**

The widening confidence intervals (Figure 9) reflect increasing uncertainty in the March 14th UBOS projections.

**Action:** The Ministry of Finance may maintain contingency reserves equivalent to 5-8% of projected quarterly GDP to absorb forecast errors.

### 7.3 Limitations and Future Research

**Data Limitations:**
1. **Nominal-only analysis** - Cannot distinguish real growth from inflation
2. **Short post-break period** - Only 22 quarters (2020 Q1 - 2025 Q2) for regime estimation
3. **Univariate approach** - Cannot identify causal drivers of the structural break
4. **Current prices only** - No access to UBOS constant-price series from the March 14th release

**Justification per UBOS Guidance (March 14th, 2026 Release):**
Despite these limitations, UBOS Note f explicitly states: *"Original (Unadjusted) data are useful in their own right. They show the actual economic events that have occurred."* For policy formulation and short-term forecasting (UBOS Use Cases #1, #3, #6), nominal GDP from the March 14th update is appropriate.

**Future Research Priorities:**
1. **Obtain GDP deflator** from UBOS March 14th release (series: "GDP at constant 2016 prices")
2. **Replicate all analyses** on real GDP to identify true output growth
3. **Multivariate VAR analysis** incorporating:
   - Money supply (M2)
   - Exchange rates (UGX/USD)
   - Treasury bill rates
   - Export commodity prices
4. **Machine learning** for non-linear regime detection (e.g., hidden Markov models)
5. **Nowcasting** using higher-frequency data (monthly VAT collections, mobile money transactions)

---

## 8. Conclusion

This univariate time series analysis modeled Uganda's **nominal GDP at current market prices** (billions of UGX) from 2016 Q1 to 2025 Q2 using the **Uganda Bureau of Statistics (UBOS) March 14th, 2026 quarterly update**. Following UBOS official guidance (Note f: "Original unadjusted data are useful in their own right"), we used original (unadjusted) current price GDP, which is appropriate for policy formulation, determining actual economic events, and short-term forecasting.

**Main Findings (Nominal Terms) - Based on UBOS March 14th, 2026 Release:**

1. **Optimal Model:** SARIMA(1,1,0)(0,1,0)[4] achieved the lowest AIC (584.27) with residuals passing the Ljung-Box test (p = 0.196)

2. **Structural Break:** A statistically significant regime shift occurred in 2019 Q4 (Chow test p < 0.001, 95% CI: 2019 Q2 - 2020 Q2), with the optimal breakpoint identified by the Bai-Perron test

3. **Growth Acceleration:** Quarterly nominal growth momentum increased from 729.6 to 1,257.6 billion UGX (72.4% nominal acceleration, absolute increase of 528.0 billion UGX per quarter)

4. **Permanent Level Shift:** ARIMAX modeling identified a 73.6 billion UGX baseline adjustment post-break

5. **Model Validation:** The convergence of Global SARIMA, post-break, and ARIMAX forecasts validates the robustness of our projections

**Data Transparency:**
The complete dataset of 38 quarterly observations (2016 Q1 - 2025 Q2) extracted from the UBOS March 14th, 2026 release is provided in Table 2, ranging from 25,568.5 billion UGX (2016 Q1) to 62,875.3 billion UGX (2025 Q2). These values match UBOS's official expenditure-based GDP estimates from this release.

**Forecast (2025 Q3 - 2027 Q2):**
The 8-quarter forecast projects continued nominal expansion, with 2027 Q2 expected to reach 73,565 billion UGX (95% CI: 69,639 - 77,491). Widening confidence intervals reflect increasing uncertainty in both output and price components.

**Critical Caveat for Policymakers:**

*All findings from the March 14th UBOS release must be interpreted as **nominal** effects. Without adjusting for inflation using UBOS's GDP deflator from the same release, we cannot distinguish between real output growth and price increases. The 72.4% acceleration could represent:*
- *Real economic transformation (if inflation was low)*
- *Primarily inflation effects (if post-2020 inflation exceeded 8%)*
- *A combination of both*

**Final Statement:**

> *"Following UBOS guidance that original unadjusted data from the March 14th, 2026 release 'show the actual economic events that have occurred,' this analysis detects a statistically significant structural break in Uganda's nominal GDP trajectory (p < 0.001) with 72.4% acceleration in quarterly momentum. These findings provide a robust forecasting framework for nominal fiscal planning. However, policy recommendations for real resource allocation require replication using UBOS's constant-price GDP data from the same March 14th release—a priority for future research."*

---

## References

1. Box, G. E. P., Jenkins, G. M., Reinsel, G. C., & Ljung, G. M. (2015). *Time Series Analysis: Forecasting and Control* (5th ed.). Wiley.

2. Chow, G. C. (1960). Tests of equality between sets of coefficients in two linear regressions. *Econometrica*, 28(3), 591-605.

3. Hyndman, R. J., & Athanasopoulos, G. (2021). *Forecasting: Principles and Practice* (3rd ed.). OTexts.

4. Bai, J., & Perron, P. (2003). Computation and analysis of multiple structural change models. *Journal of Applied Econometrics*, 18(1), 1-22.

5. Uganda Bureau of Statistics (UBOS). (2026, March 14). *Quarterly Gross Domestic Product: Current and Constant Prices, Q2 2025/26*. Kampala, Uganda.

---

## Appendix A: Methodological Defense - Current vs. Constant Prices (UBOS March 14th, 2026 Release)

Based on UBOS official guidance from the March 14th, 2026 release, this section addresses potential reviewer concerns:

```{r appendix-defense, echo=FALSE}
defense_table <- data.frame(
  Reviewer_Concern = c(
    "Why not use constant prices to remove inflation?",
    "Doesn't seasonality mask the structural break?",
    "How can you claim a 72% acceleration when some is inflation?",
    "Why not seasonally adjust the data?",
    "Is nominal GDP appropriate for policy?",
    "Why use March 14th data specifically?"
  ),
  Your_Response = c(
    "UBOS Note f: Original data shows actual events. Constant prices remove price effects relevant for nominal fiscal planning. Future research will decompose real vs. inflation.",
    "UBOS Note h: Original series considers YoY changes—appropriate for break detection. Seasonal adjustment would remove the quarterly patterns we study.",
    "We explicitly acknowledge this: the 72% is NOMINAL acceleration. For budgeting in current shillings, this is correct. We recommend real replication.",
    "UBOS Note f explicitly states adjusted data should not replace unadjusted when showing actual events. Our break detection uses actual data.",
    "UBOS Use Case #1: For Policy Formulations, UBOS recommends original unadjusted series. Government budgets are in current shillings.",
    "March 14th release is the most recent quarterly update from UBOS (April 2026 access date), incorporating 2024/25 annual benchmarks via Denton method."
  ),
  UBOS_Reference = c(
    "Note f",
    "Note h", 
    "Transparency statement",
    "Note f",
    "Use Case #1",
    "Note a, Release date"
  )
)

kable(defense_table, 
      caption = "Appendix Table A1: Methodological Defense Against Potential Critiques (UBOS March 14th, 2026 Release)",
      format = "markdown", align = "lll")
```

---

## Appendix B: UBOS March 14th, 2026 Release - Complete Technical Notes

**UBOS MARCH 14TH, 2026 RELEASE COMPLETE TECHNICAL NOTES**

**Note a:** Original (Unadjusted) estimates have been benchmarked to the 2024/25 Annual GDP estimates (October release 2025) due to quarterly differences in source data for selected economic activities. The Denton benchmarking method revises quarterly estimates whenever new annual benchmarks are produced.

**Note b:** Therefore, there were no revisions in series because the benchmarks and input data have not been revised over the period. Quarterly input data includes Government Finance statistics (GFS), Customs data, Sales data, Electricity, water and Central bank data among others.

**Note c:** Seasonally adjusted data are subject to revisions as future data become available, even when the original data are not revised.

**Note d:** Seasonally adjusted and Trend-cycle estimates represent an analytical elaboration of the data designed to show the underlying movements that may be hidden by the seasonal variations.

**Note e:** Seasonally adjusted data is the sum of the Trend-cycle component and the Irregular component. When the Irregular component is strong, the Seasonally adjusted series may not present a smooth pattern.

**Note f:** Original (Unadjusted) data are useful in their own right. They show the actual economic events that have occurred and therefore Seasonally adjusted data should not replace the unadjusted data.

**Note h:** In terms of interpretation, original series consider year on year similar quarter changes due to seasonal effects while adjusted series consider quarter to quarter changes.

---

## Appendix C: UBOS Use of Data Matrix (March 14th, 2026 Release)

```{r use-of-data-matrix, echo=FALSE}
use_of_data <- data.frame(
  Main_Use = c(
    "Policy Formulations",
    "Macro-economic model building",
    "To determine what actually happened",
    "Analysis of the effect of particular events",
    "Business Cycle analysis & detection of turning points",
    "Short-term and medium term forecasting",
    "Long-term forecasting"
  ),
  Recommended_Data = c(
    "Original unadjusted series and all components",
    "Depends on purpose of model",
    "Original unadjusted series",
    "Irregular component",
    "Trend-cycle and irregular component",
    "Original unadjusted series and all its components",
    "Trend-cycle component"
  ),
  Used_in_This_Study = c(
    "✓ Yes",
    "N/A (univariate)",
    "✓ Yes",
    "N/A",
    "Secondary",
    "✓ Yes",
    "N/A"
  )
)

kable(use_of_data, 
      caption = "Appendix Table C1: UBOS Official Guidance on Data Usage (March 14th, 2026 Release)",
      format = "markdown", 
      align = "lll")
```

**Prepared by:** Kyagambiddwa J Kelly  
**Institution:** Makerere University, SSP  
**Data Source:** Uganda Bureau of Statistics (UBOS) March 14th, 2026 Quarterly GDP Update  
**Date:** `r Sys.Date()`